# Don't re-read yourself {#sec-drry}

Imagine you open an R script you wrote a few months ago, and you cannot follow it. Well, you can read it, but it takes you a couple of goes. This chapter is about writing code you do not have to re-read. We are going to look at ways to say what you meant the first time.

::: {.overview}

## Overview {.unnumbered}

**Duration** 60 minutes

**Questions**

<!-- These map onto the sections below, in order. If you move a section, move its question. -->

- What can you do by the end of this course?
- What is wrong with writing the same line of code many times?
- How do I build a file path without typing it out?
- How do I turn working code into a function?
- What is DRRY, and how is it different from DRY?

**What you need this session**

- A session of RStudio open
- The course data. Download `fun2debug-data.zip` from
  <https://fun2debug.njtierney.com/data/> and unzip it so that a folder called
  `data` sits next to the script you are writing in. Then open that folder as
  an RStudio project, or set your working directory to it, so that
  `list.files("data/tidy")` shows you ten files.

```{r}
#| label: pkgs
#| output: false
library(countdown)
library(dplyr)
library(readr)
library(ggplot2)
library(glue)
library(here)
```

:::

## Where we are going

Somewhat like a cooking show, we are going to show you what we are making first. It is a short script, with a few functions. By the end of the course week,  you will be able to write code like this:

```{r}
#| label: where-we-are-going
#| message: false
#| fig-alt:
#|   - "Bar chart of the proportion of people studying in each age group. It is highest for 15 to 19 year olds and falls with age."
#|   - "Nine small line charts, one per age group, of the proportion studying in Tasmania from 2014 to 2023. Each line is fairly flat."
source("ch1/read_education.R")
source("ch1/plot_study.R")

education <- read_education(2014:2023)

plot_study_by(education, y = age_group)
plot_study_over_time(education, state = "Tas.")
```

The reason I think this is a worthwhile goal is that you can reason with each of these lines of code. Each line tells you what it is doing. This is sometimes called "self-documenting code".

This is education data from the [Australian Bureau of Statistics](https://www.abs.gov.au/statistics/people/education/education-and-work-australia/latest-release). It felt apt, since we had the census somewhat recently, on August 11, 2026.

To give us a direction, we are going to focus on answering this question:

**How many Australians are studying, by age group and by state, and is that changing?**

Let's talk about reading in some education data.

## Reading ten files

The data is one file per year, from 2014 to 2023.

```{r}
#| label: list-files
list.files("data/tidy")
```

Let's read one of them:

```{r}
#| label: read-one
ed_2023 <- read_csv("data/tidy/education_2023.csv")

ed_2023
```

Each row is for a given age group (`age_group`) at a given state (`state_territory`). It contains information on the number of people studying (`n_studying`), the population (`population`), and the proportion studying (`prop_studying`). The counts are in the thousands. So the first row has 22,900 people, and is not 22.9 people(!).

Notice that **year** is not in there?

That lives in the file name:

```r
ed_2023 <- read_csv("data/tidy/education_2023.csv")
```

This is pretty common. But still a bit annoying. We will get back to it.

::: {.callout-note title="Your Turn"}

If we want all 10 files read in, how should we do this?

```r
ed_2023 <- read_csv("data/tidy/education_2023.csv")
```

1. What is different between the ten file paths? What is the same?

```r
ed_2014 <- read_csv("data/tidy/education_2014.csv")
ed_2015 <- read_csv("data/tidy/education_2015.csv")
ed_2016 <- read_csv("data/tidy/education_2016.csv")
ed_2017 <- read_csv("data/tidy/education_2017.csv")
ed_2018 <- read_csv("data/tidy/education_2018.csv")
ed_2019 <- read_csv("data/tidy/education_2019.csv")
ed_2020 <- read_csv("data/tidy/education_2020.csv")
ed_2021 <- read_csv("data/tidy/education_2021.csv")
ed_2022 <- read_csv("data/tidy/education_2022.csv")
ed_2023 <- read_csv("data/tidy/education_2023.csv")
```

2. Are there any potential problems you can foresee?

There are no wrong answers!

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

:::

There are issues with code like:

```r
ed_2014 <- read_csv("data/tidy/education_2014.csv")
ed_2015 <- read_csv("data/tidy/education_2015.csv")
ed_2016 <- read_csv("data/tidy/education_2016.csv")
ed_2017 <- read_csv("data/tidy/education_2017.csv")
ed_2018 <- read_csv("data/tidy/education_2018.csv")
ed_2019 <- read_csv("data/tidy/education_2019.csv")
ed_2020 <- read_csv("data/tidy/education_2020.csv")
ed_2021 <- read_csv("data/tidy/education_2021.csv")
ed_2022 <- read_csv("data/tidy/education_2022.csv")
ed_2023 <- read_csv("data/tidy/education_2023.csv")
```

it feels a bit safe. But it is a bit repetitive! 

But what is the thing that is changing across all of these lines of code?

What if we had 50, 100, or 1,000 files like this? The typing starts to become a problem. You might make a mistake in aligning the code with the object. So we start to re-read it just to make sure we haven't made an easy to make mistake (we didn't though, but you don't know that)

### Takehomes

- Repeated code is cheap to write, but can cost you time.
- Identifying problems takes energy, even if there are no problems.

## Building the path

Let's start with the part that changes, which is the file path.

Rather than typing the whole file path out each time, can you identify the pattern? Here's the code again, with just the file path:

```r
"data/tidy/education_2014.csv"
"data/tidy/education_2015.csv"
"data/tidy/education_2016.csv"
"data/tidy/education_2017.csv"
"data/tidy/education_2018.csv"
"data/tidy/education_2019.csv"
"data/tidy/education_2020.csv"
"data/tidy/education_2021.csv"
"data/tidy/education_2022.csv"
"data/tidy/education_2023.csv"
```

We can use the `{glue}` package to make these file paths:

```{r}
#| label: glue-path
year <- 2023

glue("data/tidy/education_{year}.csv")
```

`glue()` will build it for you. Anything inside curly braces gets swapped for its value.

```{r}
year <- 2023
glue("data/tidy/education_{year}.csv")
year <- 2020
glue("data/tidy/education_{year}.csv")
```

It reads like the sentence you would say out loud.

::: {.callout-tip title="Going deeper: paste and paste0" collapse="true"}

You could also use `paste()` and `paste0()`:

```{r}
year <- 2023
paste("data/tidy/education_", year, ".csv", sep = "")

# paste, but with sep = "" (zero spaces, hence, paste0)
paste0("data/tidy/education_", year, ".csv")
```

But I think `glue` is worth learning about! You can read a bit more at a blog post I wrote, ["Glue Magic Part I"](https://www.njtierney.com/posts/2019-07-07-glue-magic-pt1/).

:::

One more thing about file paths! The path above is relative to wherever R happens to be sitting. Which can be not where you always expect.

We can use the `here()` function from `{here}` to anchor the path to where your ".RProj" is.

```{r}
#| label: here-path
glue("data/tidy/education_{year}.csv")
glue("data/tidy/education_{year}.csv") |> here()
```

::: {.callout-tip title="Read more"}

There is a longer treatment of `{here}` and project oriented workflow in R Best
Practices, at [The {here} package](https://r-best-practices.njtierney.com/project-organisation.html#the-here-package).

:::

Now pipe that into `read_csv()`.

```{r}
#| label: pipe-read
glue("data/tidy/education_{year}.csv") |> 
  here() |> 
  read_csv()
```

Notice that we have separated the file path out into its own step. But it requires specifying what year is each time:

```{r}
#| label: path-then-read
#| message: false
year <- 2023
year_path <- glue("data/tidy/education_{year}.csv") |> here()
education_2023 <- read_csv(year_path)

year <- 2022
year_path <- glue("data/tidy/education_{year}.csv") |> here()
education_2022 <- read_csv(year_path)
```

So we've sort of taken one step forward, two steps back, right now.

Let's talk about turning that part into a function.

:::: {.panel-tabset}

#### Then

```r
ed_2014 <- read_csv("data/tidy/education_2014.csv")
ed_2015 <- read_csv("data/tidy/education_2015.csv")
ed_2016 <- read_csv("data/tidy/education_2016.csv")
ed_2017 <- read_csv("data/tidy/education_2017.csv")
ed_2018 <- read_csv("data/tidy/education_2018.csv")
ed_2019 <- read_csv("data/tidy/education_2019.csv")
ed_2020 <- read_csv("data/tidy/education_2020.csv")
ed_2021 <- read_csv("data/tidy/education_2021.csv")
ed_2022 <- read_csv("data/tidy/education_2022.csv")
ed_2023 <- read_csv("data/tidy/education_2023.csv")
```

#### Now

```r
year <- 2023
year_path <- glue("data/tidy/education_{year}.csv") |> here()
ed_2023 <- read_csv(year_path)
```

::::

### Takehomes

- `glue()` builds a string by putting values inside `{ }`
- `here()` anchors a path to the top of your project
- Pulling the changing part into its own step is a start, not the finish

## Turning it into a function

A function in R has three parts:

1. A **name**. 
2. **arguments**, the things that change, and 
3. **body**, the code that runs

Let's make a function that creates the file path:

```{r}
# function name ------  # arguments: year
education_path <- function(year){
  glue("data/tidy/education_{year}.csv") |> here()
}

education_path(2023)
education_path(2022)
education_path(2021)
```

Now we can use that like so:

```{r}
education_2023 <- education_path(2023) |> read_csv()
education_2022 <- education_path(2022) |> read_csv()
education_2021 <- education_path(2021) |> read_csv()
```

Notice that `read_csv()` is repeated each time. Let's combine those together:

```{r}
#| label: read-education-one
read_education <- function(year) {
  education_path(year) |> read_csv()
}

education_2023 <- read_education(2023)
education_2022 <- read_education(2022)
education_2021 <- read_education(2021)
```

`education_path()` knows how the files are named. 

`read_education()` knows how to read a path.

This idea that we can put functions together is called: "composing" functions.

It's a good thing!

We are still doing some repetition, but bear with me here.

::: {.callout-note title="Your Turn"}

<details>
<summary>Setup code</summary>

```r
library(dplyr)
library(readr)
library(glue)
library(here)
```

</details>

1. Type out both the `education_path()` and `read_education()` functions yourself, and run `read_education(2019)`. Check you get 72 rows.
2. What happens if you ask for a year that does not exist, like 2005?
3. Are there some other ways to list file names?
4. Are there any potential issues with the way we have written this?
5. What are the benefits of composing functions like this, where they are separate?

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

:::

:::: {.panel-tabset}

#### Then

```r
ed_2014 <- read_csv("data/tidy/education_2014.csv")
ed_2015 <- read_csv("data/tidy/education_2015.csv")
ed_2016 <- read_csv("data/tidy/education_2016.csv")
ed_2017 <- read_csv("data/tidy/education_2017.csv")
ed_2018 <- read_csv("data/tidy/education_2018.csv")
ed_2019 <- read_csv("data/tidy/education_2019.csv")
ed_2020 <- read_csv("data/tidy/education_2020.csv")
ed_2021 <- read_csv("data/tidy/education_2021.csv")
ed_2022 <- read_csv("data/tidy/education_2022.csv")
ed_2023 <- read_csv("data/tidy/education_2023.csv")
```

#### Now

```r
ed_2021 <- read_education(2021)
ed_2022 <- read_education(2022)
ed_2023 <- read_education(2023)
```

::::

### Takehomes

- A function has a name, arguments, and a body
- Small functions can be composed together
- `education_path()` knows the naming, `read_education()` knows the reading

## Reading more than one year

Let's read in three years again:

```{r}
#| label: three-years
#| message: false
ed_2021 <- read_education(2021)
ed_2022 <- read_education(2022)
ed_2023 <- read_education(2023)
```

Each line is now readable. But I cannot do much with them until they are in one table.

But hang on, what is missing? There is no `year` column, so once these three sit on top of each other there is nothing to tell them apart.

We know which year we asked for, though. So we can put it back by hand, with `mutate()`:

```{r}
#| label: mutate-year
#| message: false
ed_2021 <- read_education(2021) |> mutate(year = 2021)
ed_2022 <- read_education(2022) |> mutate(year = 2022)
ed_2023 <- read_education(2023) |> mutate(year = 2023)

ed_2023
```

::: {.callout-tip title="Going deeper: where the new column lands" collapse="true"}

`mutate()` puts a new column on the far right. Often that is fine. When you want
the year up the front where you can see it, `mutate()` takes `.before` and
`.after`:

```{r}
#| label: mutate-before
#| message: false
read_education(2021) |>
  mutate(year = 2021, .before = state_territory)
```

Name the column you want it next to, rather than a position. `.before = 1` also
works, but it stops being true the moment somebody reorders the data.

:::

Now we can stack them, and every row still knows where it came from:

```{r}
#| label: bind-three
education_2021_2023 <- bind_rows(ed_2021, ed_2022, ed_2023)

education_2021_2023
```

Have a look at those three lines again, though. The year is typed twice on each one, once to read the file and once to label it.

Two copies of the same fact is exactly the thing you have to re-read to check.

So let's move it into the function, where it only has to be right once:

```{r}
#| label: read-education-mutate
#| message: false
read_education <- function(year) {
  education_path(year) |>
    read_csv() |>
    mutate(year = year)
}

read_education(2023)
```

Nobody who calls `read_education()` has to remember the labelling step now. They just get a year column.

::: {.callout-tip title="Going deeper: letting bind_rows() do the labelling" collapse="true"}

`bind_rows()` will label the pieces for you. Give it names, and `.id` writes those names into a column:

```{r}
#| label: bind-id
#| message: false
bind_rows(
  "2021" = read_csv(education_path(2021)),
  "2022" = read_csv(education_path(2022)),
  "2023" = read_csv(education_path(2023)),
  .id = "year"
)
```

One thing to notice. That `year` column is character, not a number, because names are always text.

The names have to actually reach `bind_rows()`, too. Hand it three unnamed data frames and `.id` has nothing to write, so you get 1, 2 and 3 instead of the years. It will not warn you.

This is still one line per year, though, and the year is still typed twice on each of them. We do better than that in the next section.

:::

::: {.callout-note title="Your Turn"}

<details>
<summary>Setup code</summary>

```r
library(dplyr)
library(readr)
library(glue)
library(here)

education_path <- function(year) {
  glue("data/tidy/education_{year}.csv") |> here()
}

read_education <- function(year) {
  education_path(year) |>
    read_csv() |>
    mutate(year = year)
}
```

</details>

1. Read 2014 and 2015, and bind them together.
2. What class is that `year` column? Use `class()` to check. Is it what you expected?
3. What would you have to type to do all ten years this way?

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 3)
```

<details>
<summary>Answer</summary>

```{r}
#| label: answer-two-years
#| message: false
ed_2014 <- read_education(2014)
ed_2015 <- read_education(2015)

two_years <- bind_rows(ed_2014, ed_2015)

class(two_years$year)
```

It is a number, because we typed a number into `mutate()`.

That will not stay true. The version we finish this chapter with reads the year out of the file name instead, and that arrives as text. It comes back to bite one of us in @sec-outside-in.

For step 3, ten years this way is eleven lines. Which is roughly where we came in.

</details>

:::

:::: {.panel-tabset}

#### Then

```r
ed_2014 <- read_csv("data/tidy/education_2014.csv")
ed_2015 <- read_csv("data/tidy/education_2015.csv")
ed_2016 <- read_csv("data/tidy/education_2016.csv")
ed_2017 <- read_csv("data/tidy/education_2017.csv")
ed_2018 <- read_csv("data/tidy/education_2018.csv")
ed_2019 <- read_csv("data/tidy/education_2019.csv")
ed_2020 <- read_csv("data/tidy/education_2020.csv")
ed_2021 <- read_csv("data/tidy/education_2021.csv")
ed_2022 <- read_csv("data/tidy/education_2022.csv")
ed_2023 <- read_csv("data/tidy/education_2023.csv")
```

#### Now

```r
ed_2021 <- read_education(2021)
ed_2022 <- read_education(2022)
ed_2023 <- read_education(2023)

bind_rows(ed_2021, ed_2022, ed_2023)
```

::::

### Takehomes

- The year lived in the file name, so we have to put it back deliberately
- `mutate()` will add it, one year at a time
- `bind_rows()` stacks tables on top of each other
- Updating a function means you don't need to make changes multiple times

## What read_csv already does

Before you write any more of this, let's go one step at a time.

Here is one year, as a path, and then read in:

```{r}
#| label: ladder-one
#| message: false
education_path(2021)

education_path(2021) |> read_csv()
```

Now ask for two:

```{r}
#| label: ladder-two
#| message: false
education_path(2021:2022)

two_years <- education_path(2021:2022) |> read_csv()

two_years
```

`r nrow(two_years)` rows. Two paths went in, one table came out, and I never asked for that.

So what about all ten?

```{r}
#| label: ten-paths
education_path(2014:2023)
```

Neat!

Can we hand all of them to `read_csv()`?

```{r}
#| label: vectorised
#| message: false
ten_years <- education_path(2014:2023) |> read_csv()

ten_years
```

`r nrow(ten_years)` rows. Wow!

This works for two reasons:

1. `glue()` builds ten paths when you give it ten years, and 
2. `read_csv()` will take all ten and stack them for you. 

`read_csv()` will also tell you which file each row came from, with `id`.

That argument only exists because `read_csv()` takes a vector of paths. You give
it a column name, and it stores the file path in that column.

```{r}
#| label: read-csv-id
#| message: false
education_path(2022:2023) |>
  read_csv(id = "path")
```

Notice what you get. You choose what the column is called, but not what goes in
it. It is always the path. So `id` is not a way to hand the years through, it is
a record of which file each row came from, and the year is sitting inside that
record.

So a good deal of what I was about to write by hand is already sitting in `readr`.

This is a good outcome rather than a waste of time!

Checking whether the problem is solved already is a habit worth having, and having a go yourself first is a decent way to find out what words to search for.

The year is still sitting in that path. Let's get it back out, one step at a time.

Here is a single path to work with:

```{r}
#| label: step-path
path <- education_path(2014)

path
```

`basename()` throws away the folders and keeps the file name:

```{r}
#| label: step-basename
basename(path)
```

That step matters more than it looks. `parse_number()` takes the first number it finds, and my home folder is called `nick_1`. Skip `basename()` and the year comes out as `r readr::parse_number("/Users/nick_1/data/tidy/education_2014.csv")`.

`parse_number()` pulls the number out of what is left:

```{r}
#| label: step-parse
basename(path) |> parse_number()
```

```{r}
#| label: step-character
basename(path) |> parse_number()
```

Those steps are one idea. Get the year out of a path. So let's give that idea a name:

```{r}
#| label: year-from-path
year_from_path <- function(path) {
  basename(path) |>
    parse_number()
}

year_from_path(path)
```

Now let's put all of that inside the function.

```{r}
#| label: read-education-final
#| message: false
read_education <- function(year) {
  files <- education_path(year)
  education_raw <- read_csv(files, id = "path")

  education_raw |>
    mutate(year = year_from_path(path)) |>
    select(-path)
}

education <- read_education(2014:2023)

education
```

One line now gets you every year, with the year labelled.

Three functions now, and each one knows a single thing. `education_path()` knows how the files are named, `year_from_path()` knows how to undo that, and `read_education()` knows how to read.

And notice where the fix went. It went **inside** `read_education()`, so everything that ever calls it gets the year column, and nobody who calls it had to change.

:::: {.panel-tabset}

#### Then

```r
ed_2014 <- read_csv("data/tidy/education_2014.csv")
ed_2015 <- read_csv("data/tidy/education_2015.csv")
ed_2016 <- read_csv("data/tidy/education_2016.csv")
ed_2017 <- read_csv("data/tidy/education_2017.csv")
ed_2018 <- read_csv("data/tidy/education_2018.csv")
ed_2019 <- read_csv("data/tidy/education_2019.csv")
ed_2020 <- read_csv("data/tidy/education_2020.csv")
ed_2021 <- read_csv("data/tidy/education_2021.csv")
ed_2022 <- read_csv("data/tidy/education_2022.csv")
ed_2023 <- read_csv("data/tidy/education_2023.csv")
```

#### Now

```r
education <- read_education(2014:2023)
```

::::

### Takehomes

- `read_csv()` takes a vector of paths and stacks them for you
- `read_csv(id = )` records which file each row came from
- Run `basename()` before `parse_number()`, or you parse a number out of a folder name
- A chain of steps with one job between them is a function waiting for a name
- Check whether somebody has solved it already
- When a function grows a fix, every caller gets it for free

## Don't re-read yourself

We started this hour with ten lines, and we have finished with one line that says what it does.

In programming, the term DRY comes up a bit

**DRY** stands for Don't Repeat Yourself. It says that if you find yourself copying and pasting the same code, write a function instead. Ten `read_csv()` lines, one function. DRY is right about that, and I am not arguing with it.

But I do not think repetition was ever the real problem. Repetition is a symptom.

You end up repeating yourself because you couldn't say the idea in one go. If the idea had a name, you'd have used the name. So the cause is expression, and repeated code is what unexpressed ideas look like from outside.

The version I prefer is **DRRY**: Don't Re-Read Yourself.^[Naming credit to [Miles McBain](https://milesmcbain.xyz/).]

The trigger is not "have I typed this three times". It is "will I have to read this again to work out what it does".

That fires earlier, and it fires more often. Look back at the version we finished with:

```r
education <- read_education(2014:2023)
```

You do not re-read that. You read the name, you believe it, and you get on with the analysis.

:::: {.panel-tabset}

#### Then

```r
ed_2014 <- read_csv("data/tidy/education_2014.csv")
ed_2015 <- read_csv("data/tidy/education_2015.csv")
ed_2016 <- read_csv("data/tidy/education_2016.csv")
ed_2017 <- read_csv("data/tidy/education_2017.csv")
ed_2018 <- read_csv("data/tidy/education_2018.csv")
ed_2019 <- read_csv("data/tidy/education_2019.csv")
ed_2020 <- read_csv("data/tidy/education_2020.csv")
ed_2021 <- read_csv("data/tidy/education_2021.csv")
ed_2022 <- read_csv("data/tidy/education_2022.csv")
ed_2023 <- read_csv("data/tidy/education_2023.csv")
```

#### Now

```r
source("ch1/read_education.R")

education <- read_education(2014:2023)
```

::::

### Takehomes

- DRY says do not repeat yourself, and it is right
- DRRY says do not re-read yourself, and it fires earlier and more often
- Everything DRY catches, DRRY catches too

## Summary {.unnumbered}

1. **Re-reading is the cost.** Repetition is one way to pay it, and not the only one.
2. **`glue()` and `here()`** build a path that reads like a sentence and works from anywhere.
3. **Split what changes from what reads.** `education_path()` and `read_education()` know different things.
4. **`read_csv(id = )` and `parse_number()`** put back the information that lived in the file name.
5. **Check whether it exists first.** `read_csv()` reads ten files on its own.
6. **A function is where a fix goes**, so every caller gets it without asking.

Next session we pull a function apart properly, and start doing something with all this data.
