# Don't re-read yourself {#sec-drry}

You have opened a script you wrote three months ago and not been able to follow it. I do this with my own writing too, where I read a paragraph back twice and still cannot work out what I meant. This session is about why that happens to code, and the single best thing you can do about it.

::: {.overview}

## Overview {.unnumbered}

**Duration** 60 minutes

**Questions**

<!-- These map onto the sections below, in order. If you move a section, move its question. -->

- What does the end of this course look like?
- What is wrong with writing the same line ten times?
- How do I build a file path without typing it out?
- How do I turn working code into a function?
- How do I read ten files and still know which is which?
- What is DRRY, and how is it different from DRY?

**What you need this session**

- A session of RStudio open
- The course data. Download `fun2debug-data.zip` from
  <https://fun2debug.njtierney.com/data/> and unzip it so that a folder called
  `data` sits next to the script you are writing in. Then open that folder as
  an RStudio project, or set your working directory to it, so that
  `list.files("data/tidy")` shows you ten files.

```{r}
#| label: pkgs
#| output: false
library(dplyr)
library(readr)
library(ggplot2)
library(glue)
library(here)
```

:::

## Where we are going

I want to start by showing you the end, so you know what we are working towards.

The census was on the 11th of August, so this feels like a reasonable moment to be looking at data from the [Australian Bureau of Statistics](https://www.abs.gov.au/statistics/people/education/education-and-work-australia/latest-release).

We are going to spend most of this course on one question:

**How many Australians are studying, by age group and by state, and is that changing?**

By the end of next week, the top of your analysis will look like this.

```r
source("R/read_education.R")
source("R/plot_study.R")

education <- read_education(2014:2023)

plot_study_by(education, y = age_group)
plot_study_over_time(education, state = "Tas.")
```

You can reason with each of these lines of code based on their name. The code is documenting itself.

The idea of this is to show you where we are heading. Every one of those functions is code you would have written anyway, with a name on it.

Let's talk about reading in some education data.

## Reading ten files

The data is one file per year, from 2014 to 2023.

```{r}
#| label: list-files
list.files("data/tidy")
```

Here is one of them.

```{r}
#| label: read-one
ed_2023 <- read_csv("data/tidy/education_2023.csv", show_col_types = FALSE)

ed_2023
```

Every row is one age group in one state. So the first row says that in the ACT there were 25,800 people aged 15 to 19, and 22,900 of them were studying. The columns are the state or territory, the age group, the number studying, the population of that group, and the proportion of them studying.

The counts are in thousands.

Notice what is **not** in there. Nothing in the file says which year it is. That lives in the file name, and this is extremely common in real data.

::: {.callout-note title="Your Turn"}

Before I show you what I would do, have a think about this one. There are ten files and you want all of them.

1. How would you read in these ten files? Sketch it out, or just say it out loud.
2. What is different between the ten of them, and what is the same?
3. If you wrote it the most obvious way, what would go wrong later?

There is no wrong answer here, and I am asking because the answer you give is usually the answer everybody gives, including me.

:::

Here is what I do, most of the time.

```r
ed_2014 <- read_csv("data/tidy/education_2014.csv")
ed_2015 <- read_csv("data/tidy/education_2015.csv")
ed_2016 <- read_csv("data/tidy/education_2016.csv")
ed_2017 <- read_csv("data/tidy/education_2017.csv")
ed_2018 <- read_csv("data/tidy/education_2018.csv")
ed_2019 <- read_csv("data/tidy/education_2019.csv")
ed_2020 <- read_csv("data/tidy/education_2020.csv")
ed_2021 <- read_csv("data/tidy/education_2021.csv")
ed_2022 <- read_csv("data/tidy/education_2022.csv")
ed_2023 <- read_csv("data/tidy/education_2023.csv")
```

Ten lines, four characters different between them.

The typing is not the problem, because the typing takes twenty seconds. The trade off with this is that we need to carefully read it each time, because there might be an easy to make mistake sitting in one of those lines.

And you cannot skim it. Skimming is exactly how you miss that line six says 2019 twice.

### Takehomes

- Repeated code is cheap to write and expensive to read
- The risk in a block like this is that one line is not like the others
- You cannot skim it, because skimming is how you miss that

## Building the path

Let's start with the part that changes, which is the path.

Rather than typing the whole thing out, `glue()` will build it for you. Anything inside curly braces gets swapped for its value.

```{r}
#| label: glue-path
year <- 2023

glue("data/tidy/education_{year}.csv")
```

It reads like the sentence you would say out loud, which is more than I can say for `paste0("data/tidy/education_", year, ".csv")`.

One improvement before we go on. That path is relative to wherever R happens to be sitting, and R does not always sit where you think. `here()` anchors it to the top of your project, so it works the same whichever folder you are in.

```{r}
#| label: here-path
here(glue("data/tidy/education_{year}.csv"))
```

Now pipe that into `read_csv()`.

```{r}
#| label: pipe-read
here(glue("data/tidy/education_{year}.csv")) |>
  read_csv(show_col_types = FALSE)
```

That works, and I want you to notice the shape of it. Build a path, read the path. Two jobs.

## Turning it into a function

A function in R has three parts you need right now. A **name**, which is how you call it later. Some **arguments**, which are the things that change between calls. And a **body**, which is the code that runs.

We have two jobs, so let's make two functions.

```{r}
#| label: education-path
education_path <- function(year) {
  here(glue("data/tidy/education_{year}.csv"))
}

education_path(2023)
```

```{r}
#| label: read-education-one
read_education <- function(year) {
  education_path(year) |>
    read_csv(show_col_types = FALSE)
}

read_education(2023)
```

`education_path()` knows how the files are named. `read_education()` knows how to read one. Neither of them knows the other's business.

That split is worth more than it looks. If the files get renamed next year, one function changes and the other does not care.

::: {.callout-note title="Your Turn"}

<details>
<summary>Setup code</summary>

```r
library(dplyr)
library(readr)
library(glue)
library(here)
```

</details>

1. Type out both functions yourself and run `read_education(2019)`. Check you get 72 rows.
2. What happens if you ask for a year that does not exist, like 2005?
3. Change `education_path()` so it points at `data/raw` and the `raw_education_` prefix, then run `read_education(2015)`.
4. Change it back.

Step 3 is the one to notice. You changed one function, and `read_education()` came along with you without being touched.

<details>
<summary>Answer</summary>

```r
read_education(2019)

# 2005 gives an error naming the path it tried, which is exactly what you want

education_path <- function(year) {
  here(glue("data/raw/raw_education_{year}.csv"))
}

read_education(2015)
```

`data/raw` only goes to 2019, and there is a good deal more wrong with those files than the date range. We come back to them in @sec-looking-inside.

</details>

:::

### Takehomes

- `glue()` builds a string by putting values inside `{ }`
- `here()` anchors a path to the top of your project
- A function has a name, arguments, and a body
- Two small functions beat one big one when they know different things

## Reading more than one year

One year at a time is not what we came for. Let's do three.

```{r}
#| label: three-years
ed_2021 <- read_education(2021)
ed_2022 <- read_education(2022)
ed_2023 <- read_education(2023)
```

Better than before, because each line is now readable. But we are back to one line per year, and I still cannot work with them until they are in one table.

```{r}
#| label: bind-three
bind_rows(ed_2021, ed_2022, ed_2023)
```

216 rows, which is right. And there is a problem.

**Which rows are from which year?** They are stacked together and nothing says. Look back at the columns in the file. There is no year in there, because the year was in the file name, and we have thrown the file name away.

`bind_rows()` has thought of this. Give it a named list, and `.id` writes the names into a column.

```{r}
#| label: bind-id
bind_rows(
  "2021" = ed_2021,
  "2022" = ed_2022,
  "2023" = ed_2023,
  .id = "year"
)
```

There it is. The year is back, and it came from the label rather than from the data.

::: {.callout-note title="Your Turn"}

<details>
<summary>Setup code</summary>

```r
library(dplyr)
library(readr)
library(glue)
library(here)

education_path <- function(year) {
  here(glue("data/tidy/education_{year}.csv"))
}

read_education <- function(year) {
  education_path(year) |>
    read_csv(show_col_types = FALSE)
}
```

</details>

1. Read 2014 and 2015, and bind them together with a `year` column.
2. What class is that `year` column? Use `class()` to check. Is it what you expected?
3. What would you have to type to do all ten years this way?

<details>
<summary>Answer</summary>

```r
ed_2014 <- read_education(2014)
ed_2015 <- read_education(2015)

two_years <- bind_rows("2014" = ed_2014, "2015" = ed_2015, .id = "year")

class(two_years$year)
#> [1] "character"
```

It is character, not a number, because `.id` writes the names of the list and names are always text.

Hang onto that. It will come back to bite one of us in @sec-outside-in.

For step 3, ten years this way is thirteen lines. Which is where we came in.

</details>

:::

### Takehomes

- `bind_rows()` stacks tables on top of each other
- `.id` turns the names of a list into a column
- The year lived in the file name, so we have to put it back deliberately
- Doing it by hand is still one line per year

## What read_csv already does

Before you write any more of this, try asking `read_education()` for all ten years at once.

```{r}
#| label: vectorised
read_education(2014:2023)
```

`r nrow(read_education(2014:2023))` rows. All ten years. Out of a function you wrote to read one.

That works because `glue()` builds ten paths when you give it ten years, and `read_csv()` will take all ten and stack them for you. Neither of those was my idea, and I did not have to ask for either.

`read_csv()` will also tell you where each row came from.

```{r}
#| label: read-csv-id
education_path(2022:2023) |>
  read_csv(id = "path", show_col_types = FALSE) |>
  select(path, state_territory, age_group)
```

So a good deal of what I was about to write by hand is already sitting in `readr`.

This is a good outcome rather than a wasted ten minutes. Checking whether the problem is solved already is a habit worth having, and having a go yourself first is a decent way to find out what words to search for.

Now let's put the fix inside the function.

```{r}
#| label: read-education-final
read_education <- function(year) {
  files <- education_path(year)
  names(files) <- year

  bind_rows(
    lapply(files, read_csv, show_col_types = FALSE),
    .id = "year"
  )
}

education <- read_education(2014:2023)

education
```

One line now gets you every year, with the year labelled.

And notice where the fix went. It went **inside** `read_education()`, so everything that ever calls it gets the year column, and nobody who calls it had to change.

### Takehomes

- `read_csv()` takes a vector of paths and stacks them for you
- `read_csv(id = )` records which file each row came from
- Check whether somebody has solved it already
- When a function grows a fix, every caller gets it for free

## Don't re-read yourself

We started this hour with ten lines that differed by four characters, and we have finished with one line that says what it does.

Now for the rule everyone gets told, and the rule I think is better.

**DRY** stands for Don't Repeat Yourself. It says that if you find yourself copying and pasting the same code, write a function instead. Ten `read_csv()` lines, one function. DRY is right about that, and I am not arguing with it.

But I do not think repetition was ever the real problem. Repetition is a symptom.

You end up repeating yourself because you couldn't say the idea in one go. If the idea had a name, you'd have used the name. So the cause is expression, and repeated code is what unexpressed ideas look like from outside.

The version I prefer is **DRRY**: Don't Re-Read Yourself.^[Naming credit to [Miles McBain](https://milesmcbain.xyz/).]

The trigger is not "have I typed this three times". It is "will I have to read this again to work out what it does".

That fires earlier, and it fires more often. Look back at the version we finished with:

```r
education <- read_education(2014:2023)
```

You do not re-read that. You read the name, you believe it, and you get on with the analysis. That is the whole of it.

### Takehomes

- DRY says do not repeat yourself, and it is right
- DRRY says do not re-read yourself, and it fires earlier and more often
- Everything DRY catches, DRRY catches too
- If you cannot name it, you do not have an idea yet, you have some code

## Summary {.unnumbered}

1. **Re-reading is the cost.** Repetition is one way to pay it, and not the only one.
2. **`glue()` and `here()`** build a path that reads like a sentence and works from anywhere.
3. **Split what changes from what reads.** `education_path()` and `read_education()` know different things.
4. **`bind_rows(.id = )` and `read_csv(id = )`** put back the information that lived in the file name.
5. **Check whether it exists first.** `read_csv()` reads ten files on its own.
6. **A function is where a fix goes**, so every caller gets it without asking.

Next session we pull a function apart properly, and start doing something with all this data.
