# Outside-in, inside-out {#sec-outside-in}

When writing functions, I think there are two ways general ways to write them:

1. Top Down: Write the function **as if it already exists** and I can use it, then go and define it. 
2. Bottom Up: Adapt code you've already written and you know works, into a new function.

Writing top-down code can also be called pseudocoding, although that has its own specific and special definition in computer science/algorithms.

I like Top Down, Bottom Up. But I named this chapter/idea "Outside in, Inside Out" a little while ago.

::: {.overview}

## Overview {.unnumbered}

**Duration** 60 minutes

**Questions**

<!-- These map onto the sections below, in order. If you move a section, move its question. -->

- How do I write a function before I know it works?
- How do I pull a function out of code?
- Which way should I start writing functions, and should you write one at all?
- What do I do when the first version of a function is wrong?

**What you need this session**

- A session of RStudio open, in the same folder as last time.
- Something to draw on - a pen and paper.
- `{fnmate}`, if it installed for you. Not required.

```{r}
#| label: pkgs
#| output: false
library(countdown)
library(dplyr)
library(readr)
library(ggplot2)
library(scales)
library(glue)
library(here)
```

```{r}
#| label: setup-data
#| output: false
education_path <- function(year) {
  here(glue("data/tidy/education_{year}.csv"))
}

year_from_path <- function(path) {
  basename(path) |>
    parse_number()
}

read_education <- function(year) {
  files <- education_path(year)
  education_raw <- read_csv(files, id = "path")

  education_raw |>
    mutate(year = year_from_path(path)) |>
    select(-path)
}

pct_studying <- function(data, age = "25_29") {
  data_summary <- data |>
    filter(age_group == age) |>
    summarise(
      total_studying = sum(n_studying),
      total_population = sum(population)
    ) |>
    mutate(pct = round(total_studying / total_population * 100, 1))
  data_summary
}

education <- read_education(2014:2023)

tas_2023 <- education |>
  filter(year == 2023, state_territory == "Tas.")
```

:::

## Outside-in

Outside-in means you write the function as if it already exists.

I find it really useful to imagine a workflow, or a script.

Let's discuss an example from what we've done so far.

If we knew our goal was:

> Make a plot of education numbers for each state of Australia, over time. 

And we had one input: Education Data.

I might break these steps down as:

1. Read in the data
2. Tidy up the data
3. Plot the data.

Now, I might sketch that out, with function names that match those steps:

```{r}
#| echo: true
#| eval: false

education_raw <- read_education("path/to/education.csv")

edutation_tidy <- tidy_education(education_raw)

plot_education_state(edutation_tidy)

```

A very important idea here is that this code doesn't need to work! You're just hashing out the ideas.

The point is to get used to creating functions from scratch, and thinking about them, and how they work.

For example, you might change the idea of the `read_education` data, to have two functions, one that read all the data from a directory, and one that reads a given year:

```{r}
#| echo: true
#| eval: false

education_2020_raw <- read_education_year("path/to/education.csv", year = 2020)
education_raw <- read_education("path/to/education.csv")
```

The point is to imagine the interface.

let's go back to the question we set ourselves in chapter 1:

> How many Australians are studying, by age group and by state, and is that changing?

We can do a lot of that already! The changing part is still open. So what do I actually want to type?

:::{.callout-note title="My Turn: Demo"}

Let's run through this live. Here are a few comments to guide the live-coding

```{r}
# Goal: Read in data, filter it to a given state, and plot the change in education over time

# Reading in: inputs, outputs?
# Filtering in: inputs, outputs?
# Plotting: What of, which variables over time?
  # Can I sketch out a plot by hand?
```
:::

We ended up with something like:

```r
education <- read_education(years = 2014:2023)

tas <- filter(education, state_territory == "Tas.")

plot_study_over_time(data = tas)
```

We wrote `read_education()` in chapter 1, and `filter()` is from dplyr.

I also find it really useful and interesting to sketch out a plot by hand - even if it is terrible - it helps me get a sense of what the data should be under the plot.

So outside-in has handed me exactly one thing to build. It has also told me:

- Function names
- Its inputs
- Its outputs

I didn't have to write ggplot code, and get stuck in the specifics of `aes()`, facets, theming, etc.

Those are all questions to consider! But we can think about them later. 

The thing I like about writing code like this is that it keeps you at a higher, courser level of the analysis. It's kind of like storyboarding.

### Using {fnmate}

`{fnmate}` is by [Miles McBain](https://milesmcbain.xyz), who also gave us DRRY.
It is an R package to help you rapidly template R functions. It fits alongside other "ergonomic" R packages like `{usethis}`.

It's OK if fnmate didn't install, you can see instructions on how to do that in the Setup chapter.

We can use `{fnmate}` to write the skeleton for you.

Put your cursor on a function, like `plot_study_over_time(tas)`, run the addin, and it writes this:

```r
plot_study_over_time <- function(tas) {

  NULL

}
```

I love this. It defined the function based on how you wrote the signature.

Notice:

- It named the argument `tas`, because `tas` is what was provided. I might rename that to `data`. 
- The body is `NULL`, so the function can run, but returns nothing. We didn't need to put NULL in there, but it's kind of like a signal - like when you have a page in a report that says: "This page is intentionally blank".

::: {.callout-note title="Your Turn"}

<details>
<summary>Setup code</summary>

```{r}
#| ref.label: ["pkgs", "setup-data"]
#| eval: false
#| echo: true
```

</details>

Somebody asks you:

> which state has the biggest gap in education numbers between the youngest and the oldest age groups? Is that gap closing over time?

Write some R code from this "outside in" / "top down" / pseudocode approach. You don't need to write out the entire function. Just the signature.

1. Write out the steps you need to take to answer this question.
2. Based on the steps, write the function pseudocode, top to bottom
3. For each function, write what goes in to its arguments and what comes out.
4. Do any of the functions already exist?

I find it can be really useful to sketch out plots and data structures here before you start


```r
education <- read_education(___)

___ <- ___(education, ___)

___(___)
```

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

<details>
<summary>One possible answer</summary>

> which state has the biggest gap in education numbers between the youngest and the oldest age groups? Is that gap closing over time?

1. Read education data
2. Filter data down to the youngest and oldest age groups per state
3. Pivot data wider to youngest and oldest
4. Calculate the difference between youngest and oldest, for each year
5. Plot this: X axis is time. Y axis is difference. Facet per state?

```r
education <- read_education(2014:2023)

gaps <- study_gap_by_state(education)

plot_gap_over_time(gaps)
```

It's OK if yours look different, that's fine! What we want to focus on is if you can focus on what goes in and out of each of the functions. 

`read_education()` already exists, and the other two don't.


</details>

:::


:::{.callout-tip title="Going Deeper: Practicing 'outside in' with new data" collapse="true"}

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

OK, now let's practice writing this again, but with a smaller example.

The dataset is at:

And can be downloaded with the following code:

```{r}
brolga_counts <- read_csv("https://fun2debug.njtierney.com/data/animals/brolga_year_month.csv")

brolga_counts
```

or you can go to <https://fun2debug.njtierney.com/data/> and download the brolga data there

This data contains monthly counts of the number [brolga](https://en.wikipedia.org/wiki/Brolga) seen in Australia.

There are only three columns: 

- year
- month
- n

Your question is:

> Which year was the first year to have a complete number of months of records?

1. Write out the steps to answer this question
2. Practice writing just the function signatures
3. Use fnmate to create the functions
4. Get it working!

:::


### Takehomes

- Outside-in is where you write the function signature before the function body
- This names:
  - the function, 
  - its inputs, 
  - its outputs
- Keep you at a higher level of analysis. You don't get bogged down in details
- `{fnmate}` can help you rapidly prototype functions

## Inside-out

Inside-out is the converse of "outside in".You write code, get it working, then pull/extract a function.

Let's focus on our question:

> How many Australians are studying, by age group and by state, and is that changing?

Let's build `plot_study_over_time()` by going inside-out.

Start small, with one state and one age group.

```{r}
#| label: tas-young
tas_young <- education |>
  filter(state_territory == "Tas.", age_group == "15_19")

tas_young
```

Ten rows, one per year.

:::{.callout-note title="Your Turn: Sketch the example plot"}

**Sketch the plot before you look.** Put `year` along the bottom and `n_studying` up the side. Does the line go up, down, or stay flat across the ten years?

Draw your guess, then run the code.

Your sketch doesn't need to be great! 

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 3)
```


<!-- TODO <my collapsed section with my bad drawing>. -->

:::



```{r}
#| label: fig-first-line
#| fig-alt: "A line chart of the number of 15 to 19 year olds studying in Tasmania from 2014 to 2023. The line drifts down from about 28,500 to about 24,800, wobbling on the way."
ggplot(tas_young, aes(x = year, y = n_studying)) +
  geom_line()
```

Now go back to your drawing. Does it looks sort of what you expected?

Worth a caution before you read too much into it. The number of 15 to 19 year olds in Tasmania fell over the same period, from `r format(tas_young$population[1] * 1000, big.mark = ",")` to `r format(tas_young$population[10] * 1000, big.mark = ",")`. Fewer people studying is not the same as a smaller share of them studying. Maybe we need to look at the proportion of people? 

```{r}
ggplot(tas_young, aes(x = year, y = prop_studying)) +
  geom_line()
```


### Add more lines

Let's make the plot for the other states and territories.

So drop the age filter - just filter down to each state. Then, give each age group its own little panel with `facet_wrap()`:

```{r}
#| label: fig-facets
#| fig-alt: "Nine small line charts, one per age group, of the number of people studying in Tasmania from 2014 to 2023. The 15 to 19 panel is much the highest at around 25,000, and the older panels sit near the bottom of the range."
tas <- filter(education, state_territory == "Tas.")

tas

ggplot(tas,
       aes(x = year, y = prop_studying)) +
  geom_line() +
  facet_wrap(~age_group)
```

OK, that is what I wanted. Now we can make it nicer, which can be a bit fiddly.

What changes do you think you'd like to make?

For me, these are:

- Make the lines have a colour
- Make thicker lines
- Label the y axis to have percentage
- Set the y axis to be from 0 to 100%
- Give the plot x, y, title, and subtitle
- Use the overall theme, `theme_minimal()`

```{r}
#| label: fig-polished
#| fig-alt: "The same nine panels of Tasmanian study numbers over time, now in green, with a comma separated count axis, a title, and a light theme."
ggplot(tas,
       aes(x = year, y = prop_studying)) +
  geom_line(colour = "#1B9E77", linewidth = 0.8) +
  facet_wrap(~age_group) +
  scale_y_continuous(labels = scales::label_percent(),
                     limits = c(0, 1)) +
  labs(
    x = "Year",
    y = "Studying",
    title = "Proportion of people studying, by age group",
    subtitle = "Tas."
  ) +
  theme_minimal(base_size = 12)
```

OK, so we've got a few steps here, the point was to show that going from inside-out; bottom up is a process.

### Wrapping up: adding state

Let's focus back on the rest of the question:

> How many Australians are studying, by age group and by state, and is that changing?

Let's focus on getting this per state.

Everything else in the code will work for all states, not just tasmania.

Let's write out the code for all states:

```{r}
    ggplot(education,
           aes(x = year, y = prop_studying, colour = state_territory)) +
    geom_line(linewidth = 0.8) +
    facet_wrap(~age_group) +
    scale_y_continuous(labels = label_percent(),
                       limits = c(0,1)) +
    labs(
      x = "Year",
      y = "Studying",
      title = "Proportion of people studying, by age group and state"
    ) +
    theme_minimal(base_size = 12)

```

Now, can we update this to be a function.

Name the function - what are its arguments?

```{r}
#| eval: false
#| echo: true
plot_study_over_time <- function(data){
  
}
```


Then paste in the code into the body - make sure you change the DATA!

```{r}
plot_study_over_time <- function(data){
      ggplot(data,
           aes(x = year, y = prop_studying, colour = state_territory)) +
    geom_line(linewidth = 0.8) +
    facet_wrap(~age_group) +
    scale_y_continuous(labels = label_percent(),
                       limits = c(0,1)) +
    labs(
      x = "Year",
      y = "Studying",
      title = "Proportion of people studying, by age group and state"
    ) +
    theme_minimal(base_size = 12)
}
```


```{r}
plot_study_over_time(education)
```

Remember, because we return a ggplot2 plot, we can update it with ggplot code:

```{r}
plot_study_over_time(education) + 
  labs(y = "% studying",
       colour = "State / Territory")
```


::: {.callout-note title="Your Turn"}

<details>
<summary>Setup code</summary>

```{r}
#| ref.label: ["pkgs", "setup-data"]
#| eval: false
#| echo: true
```

</details>

Here is some code that looks at the a similar question, but facets by state. Can you make this into a function where you can change the age group?

```{r}
#| label: fig-yt-working-code
#| fig-alt: "Eight small line charts, one per state and territory, of the number of 15 to 19 year olds studying from 2014 to 2023. New South Wales and Victoria dominate the range, and the smaller jurisdictions are flat near the bottom."
education |>
  filter(age_group == "15_19") |>
  ggplot(aes(x = year, y = n_studying)) +
  geom_line(colour = "#1B9E77") +
  facet_wrap(~state_territory) +
  scale_y_continuous(labels = label_number(scale = 1000, big.mark = ","))
```

Turn it into `plot_age_over_time()`.

1. **Copy.** Make an empty function and paste the code into the body. Change nothing yet.
2. **Identify.** What changes between uses? That is your argument.
3. **Abstract.** Replace the hard coded part with the argument name.
4. **Check.** You should get the same plot you started with.
5. **Default.** Give `age` a default, so `plot_age_over_time(education)` works on its own.

```r
plot_age_over_time <- function(data, ___) {
  data |>
    filter(age_group == ___) |>
    ggplot(aes(x = year, y = n_studying)) +
    geom_line(colour = "#1B9E77") +
    facet_wrap(~state_territory) +
    scale_y_continuous(labels = label_number(scale = 1000, big.mark = ","))
}
```

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 8)
```

<details>
<summary>Answer</summary>

Only open this if you have had a go.

```{r}
#| label: fig-yt-answer
#| fig-alt: "Eight small line charts, one per state and territory, of the number of 25 to 29 year olds studying from 2014 to 2023, drawn by the new function."
plot_age_over_time <- function(data, age = "15_19") {
  data |>
    filter(age_group == age) |>
    ggplot(aes(x = year, y = n_studying)) +
    geom_line(colour = "#1B9E77") +
    facet_wrap(~state_territory) +
    scale_y_continuous(labels = label_number(scale = 1000, big.mark = ",")) +
    labs(x = NULL, y = "Studying", subtitle = age)
}

plot_age_over_time(education, age = "25_29")
```

I added a subtitle while I was in there, for fun.

</details>

:::

::: {.callout-tip title="Going deeper: one more to try on your own" collapse="true"}

Here is some code I actually wrote while making this chapter. It compares the youngest and oldest age groups in a state.

```{r}
#| label: study-gap-by-hand
young <- tas_2023 |>
  filter(age_group == "15_19") |>
  summarise(pct = sum(n_studying) / sum(population) * 100)

old <- tas_2023 |>
  filter(age_group == "55_74") |>
  summarise(pct = sum(n_studying) / sum(population) * 100)

round(young$pct - old$pct, 1)
```

It works. It's also two blocks of three lines that differ by one string, and it only does Tasmania in 2023.

Turn it into a function, using the same four steps: copy, identify, abstract, iterate.

Then check you get the same answer you started with, because it's very easy to tidy something into a different result.

<details>
<summary>One possible answer</summary>

```{r}
#| label: study-gap-fn
study_gap <- function(data, young = "15_19", old = "55_74") {
  young_pct <- pct_studying(data, age = young)$pct
  old_pct <- pct_studying(data, age = old)$pct
  round(young_pct - old_pct, 1)
}

study_gap(tas_2023)
```

This one reuses `pct_studying()` from chapter 2 rather than writing the sum out twice.

If your first answer did write it out twice, that's completely normal, and noticing afterwards is the skill.

</details>

:::

### Takehomes

- Inside-out means starting from code that works, and finding the function in it
- Copy, identify what changes, abstract it into an argument, then iterate
- A warning names a symptom, and not always the cause
- A function is where a fix goes, so every caller gets it

## Which way to write functions? Do you need a function?

So which one do you reach for?

- **Outside-in** I usually write this when I'm starting a new project and want to sketch out how I think all the pieces will work together. I know that this will change a lot. It really helps me plan things out, and storyboard the ideas.
- **Inside-out** when I have something working, but I can't see the shape yet. Making a function can even faciltate exploration!

You also don't **need** to write a function.

For example ,ggplot2 often expresses itself so well, sometimes wrapping it up in a function just doesn't end up being worthwhile.

::: {.callout-note title="Your Turn"}

Think about a project of your own. 

1. Which way did you get there? Did you plan the functions, did they fall out of code you had already written?
2. Is there something you would write outside-in now.
3. Is there a function you wrote you would leave as plain code today. What did it cost you?

There are no wrong answers!

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

:::

### Takehomes

- Outside-in when you know what you want, inside-out when you have something working
- You can do both!
- You also don't have to write a function at all!

## Iteration is normal

`plot_study_over_time()` took a few goes in this chapter. This iteration is part of the process!

Nobody gets the interface right on the first shot. Well maybe you get lucky. But you typically iterate! Writing code is writing. A first draft is rough, but it gives you a starting point.

You will get faster, as you get more practice. And as you get faster, you will have greater capacity to do more, learn more. You won't need to iterate as much. But not all the time!

### Takehomes

- Writing functions is writing. The first draft might be rough, but that's normal.
- Names change as functions grow, that is progress
- Moving a line into a different function is normal work, not a mistake

## Summary {.unnumbered}

1. **Outside-in writes the function first.** Give it a name, state the inputs, outputs.
2. **Inside-out starts from working code** Copy, identify what changes, abstract it, iterate.
3. **You don't need to write a function.** 
4. **It takes iteration**. People rarely get it right the first time. Writing functions is writing.

In the next session we talk about how to open up and explore inside a function.
