# Outside-in, inside-out {#sec-outside-in}

There are two ways I to a function. Sometimes I write the function I wish existed, and then go and make it. Other times I have code that already works, and I wrap that code up into a function. This chapter does both of these approaches, and finishes with the we did in plot chapter.

::: {.overview}

## Overview {.unnumbered}

**Duration** 60 minutes

**Questions**

<!-- These map onto the sections below, in order. If you move a section, move its question. -->

- How do I write a function before I know it works?
- How do I pull a function out of code?
- Which way should I start writing functions, and should you write one at all?
- What do I do when the first version of a function is wrong?

**What you need this session**

- A session of RStudio open, in the same folder as last time
- `{fnmate}`, if it installed for you. Nothing here depends on it.

```{r}
#| label: pkgs
#| output: false
library(countdown)
library(dplyr)
library(readr)
library(ggplot2)
library(scales)
library(glue)
library(here)
```

```{r}
#| label: setup-data
#| output: false
education_path <- function(year) {
  here(glue("data/tidy/education_{year}.csv"))
}

year_from_path <- function(path) {
  basename(path) |>
    parse_number()
}

read_education <- function(year) {
  files <- education_path(year)
  education_raw <- read_csv(files, id = "path")

  education_raw |>
    mutate(year = year_from_path(path)) |>
    select(-path)
}

pct_studying <- function(data, age = "25_29") {
  data_summary <- data |>
    filter(age_group == age) |>
    summarise(
      total_studying = sum(n_studying),
      total_population = sum(population)
    ) |>
    mutate(pct = round(total_studying / total_population * 100, 1))
  data_summary
}

education <- read_education(2014:2023)

tas_2023 <- education |>
  filter(year == 2023, state_territory == "Tas.")
```

:::

## Outside-in

Outside-in means you write the function before you write its body.

You start from what you want to type, rather than what you know how to do. Here is the question we set ourselves in chapter 1:

**How many Australians are studying, by age group and by state, and is that changing?**

We can do most of that already! The changing part is still open. So what do I actually want to type?

:::{.callout-note title="My Turn: Demo"}

Let's run through this live. Here are a few comments to guide the live-coding

```{r}
# Goal: Read in data, filter it to a given state, and plot the change in education over time

# Reading in: inputs, outputs?
# Filtering in: inputs, outputs?
# Plotting: What of, which variables over time?
  # Can I sketch out a plot by hand?
```
:::

We ended up with something like:

```r
education <- read_education(years = 2014:2023)

tas <- filter(education, state_territory == "Tas.")

plot_study_over_time(data = tas)
```

We wrote `read_education()` in chapter 1, and `filter()` is from dplyr.

I also find it really useful and interesting to sketch out a plot by hand - even if it is terrible - it helps me get a sense of what the data should be under the plot.

So outside-in has handed me exactly one thing to build. It has also told me:

- Function names
- Its inputs
- Its outputs

I didn't have to write ggplot code, think about aes(), facets, colour, etc.

Those are all questions to consider - but later. Right now they don't need to happen.

The thing I like about writing code like this is that it keeps you at a higher, courser level of the analysis. It's kind of like storyboarding.

### Letting fnmate do the typing

`{fnmate}` writes the skeleton for you, straight from that line.

Put your cursor on `plot_study_over_time(tas)`, run the addin, and it writes this underneath:

```r
plot_study_over_time <- function(tas) {

  NULL

}
```

It read that line, worked out the arguments, and left you the body to fill in.

It also named the argument `tas`, because `tas` is what I passed. I would rename that to `data`. And the body is `NULL`, so the function can run, but returns nothing. We didn't need to put NULL in there, but it's kind of like a signal - like when you have a page in a big report that says: "This page is intentionally blank".

::: {.callout-important title="It's OK if fnmate didn't install"}

The {fnmate} addin saves typing. It doesn't save the thinking, and the thinking is the part we're practising. Everywhere I use it in this book I also show what it wrote, so you can type the same thing yourself.

:::

`{fnmate}` is by [Miles McBain](https://milesmcbain.xyz), who also gave us DRRY.

::: {.callout-note title="Your Turn"}

<details>
<summary>Setup code</summary>

```{r}
#| ref.label: ["pkgs", "setup-data"]
#| eval: false
#| echo: true
```

</details>

Somebody asks you which state has the biggest gap between the youngest and the oldest age groups, and whether that gap is closing.

Write the script you wish you could run. Don't write a single function.

1. Write the functions you wish existed, top to bottom, and give each one a name.
2. For each one, say what goes in and what comes out.
3. Which of them already exist?
4. Which one would you write first, and why?

```r
education <- read_education(2014:2023)

___ <- ___(education, ___)

___(___)
```

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

<details>
<summary>One possible answer</summary>

```r
education <- read_education(2014:2023)

gaps <- study_gap_by_state(education)

plot_gap_over_time(gaps)
```

Yours will look different, and that's fine. What matters is that every line has a name you could say out loud, and that you know what goes in and what comes back.

`read_education()` already exists, and the other two don't.

I would write `study_gap_by_state()` first, because `plot_gap_over_time()` can't be tested until something makes `gaps`.

</details>

:::

### Takehomes

- Outside-in means writing the function before body
- Writing it out names the function, its inputs, and what comes back
- Writing "outside-in" means you stay at a higher level of analysis, and don't get bogged down in details
- `{fnmate}` can help you rapidly prototype functions

## Inside-out

Inside-out goes the other way. You have code that works, and the function comes out of it. Or, you write code, get it working, then pull/extract a function.

Let's build `plot_study_over_time()` by going inside out or "ground up".

Start small, with one state and one age group.

```{r}
#| label: tas-young
tas_young <- education |>
  filter(state_territory == "Tas.", age_group == "15_19")

tas_young
```

Ten rows, one per year.

:::{.callout-note title="Your Turn: Sketch the example plot"}

**Sketch the plot before you look.** Put `year` along the bottom and `n_studying` up the side. Does the line go up, down, or stay flat across the ten years?

Draw your guess, then run this.

Your sketch doesn't need to be great! You can see mine below in the collapsed details section

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 3)
```


<my collapsed section with my bad drawing>.

:::



```{r}
#| label: fig-first-line
#| fig-alt: "A line chart of the number of 15 to 19 year olds studying in Tasmania from 2014 to 2023. The line drifts down from about 28,500 to about 24,800, wobbling on the way."
ggplot(tas_young, aes(x = year, y = n_studying)) +
  geom_line()
```

Now go back to your drawing. If you drew it drifting down, you were right!

It goes from `r format(tas_young$n_studying[1] * 1000, big.mark = ",")` down to `r format(tas_young$n_studying[10] * 1000, big.mark = ",")`, so that is about `r round((1 - tas_young$n_studying[10] / tas_young$n_studying[1]) * 100)` percent fewer over ten years.

Worth a caution before you read too much into it. The number of 15 to 19 year olds in Tasmania fell over the same period, from `r format(tas_young$population[1] * 1000, big.mark = ",")` to `r format(tas_young$population[10] * 1000, big.mark = ",")`. Fewer people studying is not the same as a smaller share of them studying. This plot doesn't really distinguish what you are looking at.

### Growing it

One age group is a bit thin. What about the other eight?

So drop the age filter, and give each age group its own little panel with `facet_wrap()`.

```{r}
#| label: fig-facets
#| fig-alt: "Nine small line charts, one per age group, of the number of people studying in Tasmania from 2014 to 2023. The 15 to 19 panel is much the highest at around 25,000, and the older panels sit near the bottom of the range."
tas <- filter(education, state_territory == "Tas.")

tas

tas |>
  ggplot(aes(x = year, y = n_studying)) +
  geom_line() +
  facet_wrap(~age_group)
```

That's the plot I wanted. Now it can be made nicer, which can be a bit fiddly.

The changes are:

- add a colour to the line
- Change the linewidth
- lavel the y axis numbers to have 1,000 marks
- Give the plot x, y, title, and subtitle
- Use the overall theme, `theme_minimal()`

```{r}
#| label: fig-polished
#| fig-alt: "The same nine panels of Tasmanian study numbers over time, now in green, with a comma separated count axis, a title, and a light theme."
tas |>
  ggplot(aes(x = year, y = n_studying)) +
  geom_line(colour = "#1B9E77", linewidth = 0.8) +
  facet_wrap(~age_group) +
  scale_y_continuous(labels = label_number(scale = 1000, big.mark = ",")) +
  labs(
    x = NULL,
    y = "Studying",
    title = "Proportion of people studying, by age group",
    subtitle = "Tas."
  ) +
  theme_minimal(base_size = 12)
```

OK, so we've got a few steps here, the point was to show that going from inside-out; bottom up. Is a process.

### Wrapping it up

Now ask the inside-out question. What changes between uses?

> The state. 

Everything else in that block is the same for New South Wales as it is for Tasmania.

So the state becomes an argument, and the rest becomes the body.

```{r}
#| label: fig-plot-study-over-time
#| fig-alt: "Nine panels of Tasmanian study rates from 2014 to 2023 in green, produced by a function that takes the state as an argument."
plot_study_over_time <- function(data, state) {
  data |>
    filter(state_territory == state) |>
    ggplot(aes(x = year, y = n_studying)) +
    geom_line(colour = "#1B9E77", linewidth = 0.8) +
    facet_wrap(~age_group) +
    scale_y_continuous(labels = label_number(scale = 1000, big.mark = ",")) +
    labs(
      x = NULL,
      y = "Studying",
      title = "Number of people studying, by age group",
      subtitle = state
    ) +
    theme_minimal(base_size = 12)
}

plot_study_over_time(education, state = "Tas.")
```

We moved the filter inside the plotting function.

The filter moved inside, so you hand it all ten years and name the state once.

```{r}
#| label: fig-nt
#| fig-alt: "The same nine panels for the Northern Territory. The 15 to 19 panel runs between about 6,000 and 11,500, and every panel is far smaller than the Tasmanian equivalent."
plot_study_over_time(education, state = "NT")
```

Remember, because we return a ggplot2 plot, we can update it with changes to the years, if we so desired:

```{r}
plot_study_over_time(education, state = "NT") + theme_bw()
plot_study_over_time(education, state = "NT") + labs(y = "# people studying")
```


::: {.callout-note title="Your Turn"}

<details>
<summary>Setup code</summary>

```{r}
#| ref.label: ["pkgs", "setup-data"]
#| eval: false
#| echo: true
```

</details>

Here is some code that works. It is the same idea the other way up: one age group, every state.

```{r}
#| label: fig-yt-working-code
#| fig-alt: "Eight small line charts, one per state and territory, of the number of 15 to 19 year olds studying from 2014 to 2023. New South Wales and Victoria dominate the range, and the smaller jurisdictions are flat near the bottom."
education |>
  filter(age_group == "15_19") |>
  ggplot(aes(x = year, y = n_studying)) +
  geom_line(colour = "#1B9E77") +
  facet_wrap(~state_territory) +
  scale_y_continuous(labels = label_number(scale = 1000, big.mark = ","))
```

Turn it into `plot_age_over_time()`.

1. **Copy.** Make an empty function and paste the code into the body. Change nothing yet.
2. **Identify.** What changes between uses? That is your argument.
3. **Abstract.** Replace the hard coded part with the argument name.
4. **Check.** You should get the same plot you started with.
5. **Default.** Give `age` a default, so `plot_age_over_time(education)` works on its own.

```r
plot_age_over_time <- function(data, ___) {
  data |>
    filter(age_group == ___) |>
    ggplot(aes(x = year, y = n_studying)) +
    geom_line(colour = "#1B9E77") +
    facet_wrap(~state_territory) +
    scale_y_continuous(labels = label_number(scale = 1000, big.mark = ","))
}
```

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 8)
```

<details>
<summary>Answer</summary>

Only open this if you have had a go.

```{r}
#| label: fig-yt-answer
#| fig-alt: "Eight small line charts, one per state and territory, of the number of 25 to 29 year olds studying from 2014 to 2023, drawn by the new function."
plot_age_over_time <- function(data, age = "15_19") {
  data |>
    filter(age_group == age) |>
    ggplot(aes(x = year, y = n_studying)) +
    geom_line(colour = "#1B9E77") +
    facet_wrap(~state_territory) +
    scale_y_continuous(labels = label_number(scale = 1000, big.mark = ",")) +
    labs(x = NULL, y = "Studying", subtitle = age)
}

plot_age_over_time(education, age = "25_29")
```

I added a subtitle while I was in there, because a plot of one age group should say which one.

</details>

:::

::: {.callout-tip title="Going deeper: one more to try on your own" collapse="true"}

Here is some code I actually wrote while making this chapter. It compares the youngest and oldest age groups in a state.

```{r}
#| label: study-gap-by-hand
young <- tas_2023 |>
  filter(age_group == "15_19") |>
  summarise(pct = sum(n_studying) / sum(population) * 100)

old <- tas_2023 |>
  filter(age_group == "55_74") |>
  summarise(pct = sum(n_studying) / sum(population) * 100)

round(young$pct - old$pct, 1)
```

It works. It's also two blocks of three lines that differ by one string, and it only does Tasmania in 2023.

Turn it into a function, using the same four steps: copy, identify, abstract, iterate.

Then check you get the same answer you started with, because it's very easy to tidy something into a different result.

<details>
<summary>One possible answer</summary>

```{r}
#| label: study-gap-fn
study_gap <- function(data, young = "15_19", old = "55_74") {
  young_pct <- pct_studying(data, age = young)$pct
  old_pct <- pct_studying(data, age = old)$pct
  round(young_pct - old_pct, 1)
}

study_gap(tas_2023)
```

This one reuses `pct_studying()` from chapter 2 rather than writing the sum out twice.

If your first answer did write it out twice, that's completely normal, and noticing afterwards is the skill.

</details>

:::

### Takehomes

- Inside-out means starting from code that works, and finding the function in it
- Copy, identify what changes, abstract it into an argument, then iterate
- A warning names a symptom, and not always the cause
- A function is where a fix goes, so every caller gets it

## Which way, and whether to bother

So which one do you reach for?

- **Outside-in** when I know what I want and not how. Planning, drafting, the top of a script, the moment somebody asks me a question. You can also think of this as "sketching the analysis out" or "storyboarding".
- **Inside-out** when I have something that works and cannot see the shape yet. Exploring, poking, the second time I paste something.

You also don't **need** to write a function.

Chapter 2 showed this with a plain ggplot call. ggplot2 expresses itself well, so wrapping a single call in a function buys you a name and not much else.
::: {.callout-note title="Your Turn"}

Think about a project of your own. No code for this one.

1. Which way did you get there? Did you plan the functions, or did they fall out of code you had already written?
2. Name one thing in it you would write outside-in now.
3. Name one function you wrote that you would leave as plain code today. What did it cost you?

There are no wrong answers, and I would like to hear number 3 especially.

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

:::

### Takehomes

- Outside-in when you know what you want, inside-out when you have something working
- Most people do both, often on the same script
- Not writing a function is a real answer
- A function costs a name, a definition, and an interface you have to keep

## Iteration is normal

`plot_study_over_time()` took four goes in this chapter, and the first one drew nothing.

That's not a teaching device. That's what it looks like.

The interface moved too. Here is the line I wrote outside-in at the top of the chapter, and the one we finished with:

```r
plot_study_over_time(tas)

plot_study_over_time(education, state = "Tas.")
```

The first one keeps the filter outside, and the last one pulls it in and grows an argument to do it. The process requires iteration: I only worked out which one I wanted by writing the body.

Nobody gets the interface right the first time around. Writing code is writing. A first draft is rough, but it also means there is something to improve.

As you get more practice writing code, and writing functions, you will iterate faster. As you iterate faster, you will learn more, and start requiring as many iterations. But not always! 

### Takehomes

- The first version is a draft, and drafts are meant to be rough
- Names change as functions grow, and that is progress
- Moving a line into a different function is normal work, not a mistake

## Summary {.unnumbered}

1. **Outside-in writes the function first, and the body second.** Give it a name, say what goes in and what comes back, then make it true.
2. **Inside-out starts from code that works.** Copy, identify what changes, abstract it, iterate.
3. **Not writing a function is a real answer.** Clear, used once, and hard to name are all reasons to leave it.
4. **Nobody gets it right first go.** Four versions is normal, and renaming is part of writing.

Next session we open a function up and look inside it while it runs.
